---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# calamineR

Fast Excel file reader for R, powered by the Rust [calamine](https://github.com/tafia/calamine) library.

## Supported Formats

- xlsx (Excel 2007+)
- xlsm (Excel with macros)
- xlsb (Excel Binary)
- xls (Excel 97-2003)
- ods (OpenDocument Spreadsheet)

## System Requirements

- R >= 4.0
- Rust toolchain (cargo >= 1.75.0) - install from https://rustup.rs/

## Installation

You can install the development version of calamineR from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("MaxManek1110/calamineR")
```

## Usage

```{r example, eval = FALSE}
library(calamineR)

# Read a sheet as data.frame
df <- read_excel("data.xlsx")
df <- read_excel("data.xlsb", sheet = "Sheet2")
df <- read_excel("data.xlsx", sheet = 2, col_names = FALSE)
df <- read_excel("data.xlsx", skip = 5)  # Skip first 5 rows

# Fill merged cells with the top-left cell value
df <- read_excel("data.xlsx", fill_merged_cells = TRUE)

# Get sheet names
sheets <- excel_sheets("data.xlsx")

# Get sheet dimensions
dims <- sheet_dims("data.xlsx", 1)
# dims["rows"], dims["cols"]

# Read as raw list of rows (for complex layouts)
rows <- read_sheet_raw("data.xlsx", "Sheet1")
```

## Functions

| Function | Description |
|----------|-------------|
| `read_excel()` | Read sheet as data.frame |
| `excel_sheets()` | Get sheet names |
| `sheet_dims()` | Get sheet dimensions (rows, cols) |
| `read_sheet_raw()` | Read as list of character vectors |

## Parameters for `read_excel()`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `path` | character | (required) | Path to the Excel file |
| `sheet` | character/integer | `1L` | Sheet name or 1-based index |
| `col_names` | logical | `TRUE` | Use first row as column names |
| `skip` | integer | `0L` | Number of rows to skip |
| `fill_merged_cells` | logical | `FALSE` | Fill merged cells with top-left value |

## Merged Cells Support

When `fill_merged_cells = TRUE`, cells that are part of a merged region are filled with the value from the top-left cell of that region. This is useful when reading spreadsheets where headers or data span multiple cells.

Supported formats: xlsx, xlsm, xlsb, xls (ods not yet supported for merged cells).

## Performance

Calamine is written in pure Rust and is significantly faster than many alternatives:

- 1.75x faster than excelize (Go)
- 7x faster than ClosedXML (C#)
- 9x faster than openpyxl (Python)

## Benchmark

Comparison against `readxlsb` for reading a large randomized Excel file (~40MB)
As it is not possible to run benchmarks on CRAN, the following results are based on
local testing and may vary based on hardware and file structure.

Summary: 

- calamineR: 3.438 s
- readxlsb:  451.717 s
- 131.4 x faster

```{r benchmark, eval = FALSE}
library(calamineR)
library(readxlsb)

# Take a large randomized Excel file from Microsoft directly (~40MB)
file_path <- "inst/extdata/random_table_40mb.xlsb"


time_calamineR <- system.time({
  df_calamineR <- calamineR::read_excel(file_path)
})
sheets <- calamineR::excel_sheets(file_path)
dims <- calamineR::sheet_dims(file_path, sheet = 1)
range <- paste0("A1:", LETTERS[dims[2][[1]]], dims[1][[1]])
cat("  Time:", time_calamineR["elapsed"], "seconds\n")
cat("  Rows:", nrow(df_calamineR), " Cols:", ncol(df_calamineR), "\n\n")

cat("readxlsb::read_xlsb()\n")
time_readxlsb <- system.time({
  df_readxlsb <- readxlsb::read_xlsb(
    file_path,
    sheet = sheets[1],
    range = range
  )
})

# Summary
cat("--- Summary ---\n")
cat("calamineR:", round(time_calamineR["elapsed"], 3), "s\n")
cat("readxlsb: ", round(time_readxlsb["elapsed"], 3), "s\n")
cat(
  "Speedup:  ",
  round(time_readxlsb["elapsed"] / time_calamineR["elapsed"], 1),
  "x faster\n"
)

identical(df_readxlsb, df_calamineR)
```

## Development

`calamineR` is built using `extendr` and `Rust`.
To build the package, you need to have `Rust` installed on your machine.

```r
rextendr::document()
devtools::document()
```
